Appendix 6: sdmTMB speed testing methods and model validation description

### 2022-01-28 doc/appendix-binomial-inla.Rmd
Edits dealing with Jim and Joe's comments

```{r setup, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  cache = TRUE,
  autodep = TRUE
)
```

@@

Note that the standard errors for variance terms are not in natural and not log space and so are a very rough approximation and therefore not reported by the `tidy()` function.

>>>>

```{r setup, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  cache = TRUE,
  autodep = TRUE,
  warning = FALSE,
  message = FALSE
)
```

@@

Note that, within sdmTMB, the standard errors for variance terms are not reported in the `tidy()` function as they could be misleading if considered in natural (rather than log) space.
The confidence intervals are, however, reported.

### 2022-01-28 doc/appendix-model.Rmd
Edits dealing with Jim and Joe's comments

### Smoothers
<!-- EW: see this paragraph and below, should package names be enclosed in single quotes, like `sdmTMB()` above? Examples: mgcv, brms, sdmTMB throughout -->
<!-- EW: unrelated, I never tried to implement factor smooths. I assume they work, but it'd good to check eventually -->
Smoothers in sdmTMB are implemented with the same formula syntax familiar to mgcv users fitting GAMs (generalized additive models).
Smooths are implemented in the formula using `+ s(x)`, which implements a smooth from `mgcv::s()`.
Within these smooths, the same syntax commonly used in `mgcv::s()` can be applied, e.g. 2-dimensional smooths may be constructed with `+ s(x, y)`; smooths can be specific to various factor levels, `+ s(x, by = group)`; smooths can vary according to a continuous variable, `+ s(x, by = x2)`; the basis function dimensions may be specified, e.g. `+ s(x, k = 4)`; and various types of splines may be constructed such as cyclic splines to model seasonality, e.g. `+ s(month, bs = "cc", k = 12)`.

@@

where $\tau^2$ controls the spatial variance,
$\nu$ controls the smoothness,
$\Gamma$ represents the Gamma function,
$d_{jk}$ represents the distance between locations $s_j$ and $s_k$,
$K_\nu$ represents the modified Bessel function of the second kind,
and $\kappa$ controls the spatial correlation range (decorrelation rate).
The parameter $\nu$ is set to 1 to take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GMRF to greatly increase computational efficiency [@lindgren2011].
Internally, the parameter $\kappa$ and $\tau$ are converted to range and marginal standard deviation $\sigma$ as $\textrm{range} = \sqrt{8} / \kappa$ and $\sigma = 1 / \sqrt{4 \pi \exp \left(2 \log(\tau) + 2 \log(\kappa) \right) }$.

@@

## Barrier mesh

In some cases the spatial domain of interest may be complex and bounded by some barrier such as by land or water (e.g., coastlines, islands, lakes).
SPDE models allow for barriers to be included via barrier meshes [@bakka2019].
With  `sdmTMB()` models, the mesh construction occurs in two steps: the user (1) constructs a mesh with a call to `sdmTMB::make_mesh()`, and (2) passes the mesh to `sdmTMB::add_barrier_mesh()`.
The barriers must be constructed as `sf` objects [@pebesma2018] with polygons defining the barriers.
See `?sdmTMB::add_barrier_mesh` for examples.

The barrier mesh implementation requires the user to select a fraction value (`range_fraction` argument) that defines the fraction of the usual spatial range when crossing the barrier [@bakka2019].
For example, if the range was estimated at 10 km, `range_fraction = 0.2` would assume that the range was 2 km across the barrier.
This would let the spatial correlation decay 5 times faster with distance.
From experimentation, values around 0.1 or 0.2 seem to work well but values much lower than 0.1 can result in convergence issues.

>>>>

### Smoothers

Smoothers in sdmTMB are implemented with the same formula syntax familiar to mgcv users fitting GAMs (generalized additive models).
Smooths are implemented in the formula using `+ s(x)`, which implements a smooth from `mgcv::s()`.
Within these smooths, the same syntax commonly used in `mgcv::s()` can be applied, e.g. 2-dimensional smooths may be constructed with `+ s(x, y)`; smooths can be specific to various factor levels, `+ s(x, by = group)`; smooths can vary according to a continuous variable, `+ s(x, by = x2)`; the basis function dimensions may be specified, e.g. `+ s(x, k = 4)`; and various types of splines may be constructed such as cyclic splines to model seasonality, e.g. `+ s(month, bs = "cc", k = 12)`.

@@

where $\tau^2$ controls the spatial variance,
$\nu$ controls the smoothness,
$\Gamma$ represents the Gamma function,
$d_{jk}$ represents the distance between locations $s_j$ and $s_k$,
$K_\nu$ represents the modified Bessel function of the second kind,
and $\kappa$ controls the spatial correlation range (decorrelation rate).
The parameter $\nu$ is set to 1 to take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GRF to greatly increase computational efficiency [@lindgren2011].
Internally, the parameter $\kappa$ and $\tau$ are converted to range and marginal standard deviation $\sigma$ as $\textrm{range} = \sqrt{8} / \kappa$ and $\sigma = 1 / \sqrt{4 \pi \exp \left(2 \log(\tau) + 2 \log(\kappa) \right) }$.

@@

## Incorporating physical barriers into the SPDE

In some cases the spatial domain of interest may be complex and bounded by some barrier such as by land or water (e.g., coastlines, islands, lakes).
SPDE models allow for physical barriers to be incorporated into the modelling [@bakka2019].
With  `sdmTMB()` models, the mesh construction occurs in two steps: the user (1) constructs a mesh with a call to `sdmTMB::make_mesh()`, and (2) passes the mesh to `sdmTMB::add_barrier_mesh()`.
The barriers must be constructed as `sf` objects [@pebesma2018] with polygons defining the barriers.
See `?sdmTMB::add_barrier_mesh` for examples.

The barrier implementation requires the user to select a fraction value (`range_fraction` argument) that defines the fraction of the usual spatial range when crossing the barrier [@bakka2019].
For example, if the range was estimated at 10 km, `range_fraction = 0.2` would assume that the range was 2 km across the barrier.
This would let the spatial correlation decay 5 times faster with distance.
From experimentation, values around 0.1 or 0.2 seem to work well but values much lower than 0.1 can result in convergence issues.

### 2022-01-28 doc/appendix-pcod-VAST-tweedie.Rmd
avoid conflict between versions of make_mesh in VAST appendix
Edits dealing with Jim and Joe's comments

Unlike in sdmTMB, the fitting and predicting steps are all accomplished with the function 'fit_model' and thus we need to specify the prediction grid (referred to as the "extrapolation grid" in VAST). Several options are built in for commonly used regions, but in this example we will need to format the Queen Charlotte Sound grid in sdmTMB to be used in VAST and specify `Region = "User"` in `make_setting()` and `"input_grid" = input_grid` in `fit_model()`.

@@

```{r sdmTMB-example, echo = TRUE, cache=TRUE}
mesh <- make_mesh(pcod, xy_cols = c("X", "Y"),
  mesh = fit$spatial_list$MeshList$isotropic_mesh)

>>>>

Unlike in sdmTMB, the fitting and predicting steps are all accomplished with the function 'fit_model' and thus we need to specify the prediction grid (referred to as the "extrapolation grid" in VAST). Several options are built in for commonly used regions, but in this example we will need to format the Queen Charlotte Sound grid in sdmTMB to be used in VAST and specify `Region = "User"` in `make_setting()` and `input_grid = input_grid` in `fit_model()`.

@@

```{r sdmTMB-example, echo = TRUE, cache=TRUE}
mesh <- sdmTMB::make_mesh(pcod, xy_cols = c("X", "Y"), 
                          mesh = fit$spatial_list$MeshList$isotropic_mesh)

### 2022-01-28 doc/appendix-validation.Rmd
Edits dealing with Jim and Joe's comments

```{r mgcv-eg1, echo=TRUE}
set.seed(19203)
# examples from ?mgcv::gam.models
dat <- mgcv::gamSim(3, n = 800)
m_mgcv <- mgcv::gam(y ~ s(x2), data = dat)
p_mgcv <- predict(m_mgcv)
dat$X <- runif(nrow(dat)) # fake
dat$Y <- runif(nrow(dat)) # fake
spde <- make_mesh(dat, c("X", "Y"), cutoff = 0.1)
m <- sdmTMB(y ~ s(x2),
  data = dat,
  mesh = spde, spatial = "off", spatiotemporal = "off"
)
p <- predict(m, newdata = NULL)
plot(p$est, p_mgcv)
abline(a = 0, b = 1)
cor(p$est, p_mgcv)
```

We can see the predictions match nearly exactly.
Furthermore, we can use the in-development function `sdmTMB::plot_smooth()` to compared the smoothers themselves.
Note that the sdmTMB version is showing the marginal effect of a prediction including the intercept whereas the mgcv version is showing only the smooth component.
Nonetheless, here the smoothers look very similar except for these scale on the y axis.

```{r plot-smooth, echo=TRUE, fig.asp=0.6, out.width="5in"}
plot_smooth(m)
plot(m_mgcv)
```

@@

```{r bivariate-compare, echo=TRUE}
m_mgcv <- mgcv::gam(y ~ s(x2, x1), data = dat)
p_mgcv <- predict(m_mgcv)
m <- sdmTMB(y ~ s(x2, x1),
  data = dat,
  mesh = spde, spatial = "off", spatiotemporal = "off"
)
p <- predict(m, newdata = NULL)
plot(p$est, p_mgcv)
abline(a = 0, b = 1)
cor(p$est, p_mgcv)
```

>>>>

```{r mgcv-eg1, echo=TRUE}
set.seed(19203)
# examples from ?mgcv::gam.models
dat <- mgcv::gamSim(3, n = 800)
m_mgcv <- mgcv::gam(y ~ s(x2), data = dat, method = "REML")
p_mgcv <- predict(m_mgcv)
dat$X <- runif(nrow(dat)) # fake
dat$Y <- runif(nrow(dat)) # fake
spde <- make_mesh(dat, c("X", "Y"), cutoff = 0.1)
m <- sdmTMB(y ~ s(x2),
  data = dat,
  mesh = spde, spatial = "off", spatiotemporal = "off"
)
p <- predict(m, newdata = NULL)
plot(p$est, p_mgcv)
abline(a = 0, b = 1)
cor(p$est, p_mgcv)
```

We can see the predictions match nearly exactly.
Furthermore, we can use the in-development function `sdmTMB::plot_smooth()` to compare the smoothers themselves.
Note that the sdmTMB version shows the marginal effect of a prediction including the intercept whereas the `mgcv::plot.gam()` shows the smooth component only.
We therefore use `mgcv::predict.gam()` to make a matching plot.
Below, we add a tiny jitter amount (`0.02`) to the mgcv output just so the two smoothers can both be visualized.
The blue lines are from mgcv and the black lines are from sdmTMB.

```{r plot-smooth, echo=TRUE, fig.asp=0.6, out.width="5in"}
plot_smooth(m)

x2 <- seq(0, 1, length.out = 100)
p_mgcv <- predict(m_mgcv, newdata = data.frame(x2 = x2), se.fit = TRUE)
qv <- stats::qnorm(1 - (1 - 0.95) / 2)
df <- data.frame(
  x2 = x2,
  est = p_mgcv$fit, 
  lwr = p_mgcv$fit - qv * p_mgcv$se.fit, 
  upr = p_mgcv$fit + qv * p_mgcv$se.fit
)
lines(df$x2, df$est + 0.02, col = "blue")
lines(df$x2, df$lwr + 0.02, lty = 2, col = "blue")
lines(df$x2, df$upr + 0.02, lty = 2, col = "blue")
```

@@

```{r bivariate-compare, echo=TRUE}
m_mgcv <- mgcv::gam(y ~ s(x2, x1), data = dat, method = "REML")
p_mgcv <- predict(m_mgcv)
m <- sdmTMB(y ~ s(x2, x1),
  data = dat,
  mesh = spde, spatial = "off", spatiotemporal = "off"
)
p <- predict(m, newdata = NULL)
plot(p$est, p_mgcv)
abline(a = 0, b = 1)
cor(p$est, p_mgcv)
```
